---
title: "SpatialScope Function-Level Analysis Guide"
author: "Mengyao Lu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpatialScope Function-Level Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

SpatialScope is an interactive platform for region-specific analysis and visualization of spatial transcriptomics data. While the full GUI can be launched using `run_SpatialScope()`, the package also provides a set of individual functions that enable reproducible, programmatic workflows.

This vignette demonstrates how these core functions can be combined to build customized spatial analysis pipelines.

## Installation
```{r eval=FALSE}
# Install from GitHub
devtools::install_github("myaol/SpatialScope")
```

## Load Library
```{r eval=FALSE}
library(SpatialScope)
```

# Region of Interest (ROI) Selection

## Interactive ROI Drawing

The `draw_ROI()` function enables interactive selection of spatial regions:
```{r eval=FALSE}
# Launch interactive ROI selection tool
Seurat_object <- readRDS("path/to/your_seurat.rds")

# Or load the built-in demo dataset
demo_path <- system.file("extdata", "example_visium.rds", package = "SpatialScopeDev")
Seurat_object <- readRDS(demo_path)

# Draw ROI and return selected spot IDs
selected_spots <- draw_ROI(Seurat_object, sample_name = "MyExperiment")

# Create subset of data based on selected region
subset_Seurat_object <- subset(Seurat_object, cells = selected_spots)
```

**Key Features:**
- Click-and-drag to select regions of interest
- Returns spot IDs for downstream analysis
- Enables region-specific analyses

# Interactive Viewers

SpatialScope provides several interactive viewers that can be launched with single function calls:

## Gene Viewer

Visualize individual gene expression patterns across spatial data:
```{r eval=FALSE}
# Launch interactive gene viewer
run_gene_viewer(Seurat_object)

# Or use on subset data
run_gene_viewer(subset_Seurat_object)
```

The gene viewer provides:
- Interactive spatial visualization of gene expression
- Search functionality for genes in the dataset
- Real-time expression pattern exploration

## Gene Set Scoring Viewer

Analyze gene set enrichment patterns spatially:
```{r eval=FALSE}
# Launch gene set scoring interface
run_gene_set_viewer(Seurat_object, sample_name = "MyExperiment")

# Or use on subset data
run_gene_set_viewer(subset_Seurat_object, sample_name = "MyExperiment")
```

This interface allows to:
- Define custom gene sets
- Score spots based on gene set expression
- Visualize gene set enrichment spatially

# Spatial Clustering Analysis

## Running Spatial Clustering

The `run_spatial_clustering()` function performs graph-based clustering on spatial data:
```{r eval=FALSE}
# Perform spatial clustering on ROI
result <- run_spatial_clustering(
  subset_Seurat_object,
  resolution = 0.8,    # Range: 0.3-2.0 (higher = more clusters)
  n_pcs = 30,          # Range: 5-50 (number of principal components)
  run_umap = TRUE,
  plot_umap = TRUE,
  plot_spatial = TRUE,
  verbose = TRUE
)
```


**Outputs:**
The function generates:
- UMAP visualization 
- Spatial cluster plot 
- Console output with clustering statistics
- Modified Seurat object with cluster assignments

## Extracting Cluster Information
```{r eval=FALSE}
# Extract spot IDs for specific cluster(s)
cluster_0_spots <- get_cluster_spots(result, clusters = "0")

# Extract multiple clusters
cluster_0_1_spots <- get_cluster_spots(result, clusters = c("0", "1"))
```

# Differential Expression Analysis

## Running DEG Analysis

The `run_deg_analysis()` function identifies differentially expressed genes between spatial regions:
```{r eval=FALSE}
# Compare selected region vs rest of tissue
degs <- run_deg_analysis(
  Seurat_object,
  group1_spots = selected_spots,
  comparison = "group1_vs_rest"
)
```

**Comparison Types:**
- `"group1_vs_rest"`: Compare selected spots against all other spots
- `"group1_vs_group2"`: Compare two specific groups 

## Examining DEG Results
```{r eval=FALSE}
# View top 20 differentially expressed genes
head(degs, 20)

# Check all available columns
colnames(degs)
# Returns: gene, avg_log2FC, pct.1, pct.2, p_val, p_val_adj

# Save complete results
write.csv(degs, "DEG_results_all.csv", row.names = FALSE)
```

**Result Columns:**
- `gene`: Gene name
- `avg_log2FC`: Average log2 fold change
- `pct.1`: Percentage of cells expressing in group 1
- `pct.2`: Percentage of cells expressing in group 2
- `p_val`: Unadjusted p-value
- `p_val_adj`: Adjusted p-value (for multiple testing)

Results are automatically ranked by adjusted p-value.

# Feature Comparison and Visualization

## Group Comparison Plot

Compare expression patterns between two spatial groups:
```{r eval=FALSE}
# Compare gene expression between two groups
plot_group_comparison(
  Seurat_object,
  feature = "PIGR",                          # Gene name
  feature_type = "gene",                      # "gene", "metadata", or "geneset"
  group1_spots = group1_spots,
  group2_spots = group2_spots,
  comparison = "group1_vs_group2",            # Comparison type
  stat_test = "wilcox"                        # Statistical test
)
```

**Parameters:**
- `feature`: Name of feature to compare (gene, metadata column, or gene set)
- `feature_type`: Type of feature
  - `"gene"`: Individual gene expression
  - `"metadata"`: Metadata column (e.g., cluster assignment)
  - `"geneset"`: Gene set score
- `comparison`: Comparison type
  - `"group1_vs_group2"`: Compare two specific groups
  - `"group1_vs_rest"`: Compare group 1 against all other spots
  - `"group2_vs_rest"`: Compare group 2 against all other spots
- `stat_test`: Statistical test to use
  - `"wilcox"`: Wilcoxon rank-sum test (non-parametric)
  - `"ttest"`: Student's t-test (parametric)

## Feature-to-Feature Comparison

Compare two features within a specific group of spots:
```{r eval=FALSE}
# Compare expression of two genes within selected region
plot_feature_violin_comparison(
  Seurat_object,
  feature1 = "PIGR",
  feature2 = "MT1G",
  feature1_type = "gene",
  feature2_type = "gene",
  spots_to_use = group1_spots,                # Analyze within specific spots
  stat_test = "wilcox"                        # Wilcoxon test
)
```

**Use Cases:**
- Compare expression of two genes in same region
- Compare two gene set scores
- Identify co-expression patterns

# Complete Analysis Workflow Example

Here's a complete analysis combining multiple SpatialScope functions:
```{r eval=FALSE}
library(SpatialScope)

# Step 1: Load data
Seurat_object <- readRDS("path/to/Seurat_object.rds")

# Step 2: Select regions of interest
selected_spots_group1 <- draw_ROI(Seurat_object, sample_name = "MyExperiment")
selected_spots_group2 <- draw_ROI(Seurat_object, sample_name = "MyExperiment")

# Step 3: Subset data for each group
group1_subset_Seurat_object <- subset(Seurat_object, cells = selected_spots_group1)
group2_subset_Seurat_object <- subset(Seurat_object, cells = selected_spots_group2)

# Step 4: Explore genes interactively
# Can explore the whole dataset
run_gene_viewer(Seurat_object)

# Or explore specific subset
run_gene_viewer(group1_subset_Seurat_object)

# Step 5: Score gene sets
# On whole dataset
run_gene_set_viewer(Seurat_object, sample_name = "Whole_Analysis")

# Or on subset
run_gene_set_viewer(group1_subset_Seurat_object, sample_name = "Group1_Analysis")

# Step 6: Cluster group1 region
group1_cluster <- run_spatial_clustering(
  group1_subset_Seurat_object,
  resolution = 0.8,
  n_pcs = 30,
  run_umap = TRUE,
  plot_umap = TRUE,
  plot_spatial = TRUE
)

# Step 7: Extract specific cluster
cluster_2_spots <- get_cluster_spots(group1_cluster, clusters = "2")

# Step 8: Find differentially expressed genes
# Compare group1 vs group2
group1_vs_group2_degs <- run_deg_analysis(
  Seurat_object,
  group1_spots = selected_spots_group1,
  group2_spots = selected_spots_group2,
  comparison = "group1_vs_group2"
)

# Compare specific cluster vs rest of group1
cluster_2_degs <- run_deg_analysis(
  group1_subset_Seurat_object,
  group1_spots = cluster_2_spots,
  comparison = "group1_vs_rest"
)

# Step 9: Visualize top gene between groups
top_gene <- group1_vs_group2_degs$gene[1]
plot_group_comparison(
  Seurat_object,
  feature = top_gene,
  feature_type = "gene",
  group1_spots = selected_spots_group1,
  group2_spots = selected_spots_group2,
  comparison = "group1_vs_group2",
  stat_test = "wilcox"
)

# Step 10: Compare two marker genes (feature-to-feature comparison)
plot_feature_violin_comparison(
  Seurat_object,
  feature1 = group1_vs_group2_degs$gene[1],
  feature2 = group1_vs_group2_degs$gene[2],
  feature1_type = "gene",
  feature2_type = "gene",
  spots_to_use = selected_spots_group1,
  stat_test = "wilcox"
)

# Step 11: Export results
write.csv(group1_vs_group2_degs, "group1_vs_group2_DEGs.csv", row.names = FALSE)
write.csv(cluster_2_degs, "cluster_2_DEGs.csv", row.names = FALSE)
```



## Data Management
```{r eval=FALSE}
# Save spot selections for reproducibility
saveRDS(selected_spots_group1, "group1_roi_spots.rds")
saveRDS(selected_spots_group2, "group2_roi_spots.rds")

# Save clustered object
saveRDS(group1_cluster, "group1_clustered_data.rds")
```

